Bayesian RKHS-based methods in functional regression\support{Support information of the article.}

### 2022-02-25 paper/paper.tex
Update paper

\newcommand{\N}{\mathbb{N}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\endlocaldefs

@@

%% Inline comments
\newcommand\incomment[1]{\color{red}[\textit{#1}]\color{black}}

@@

@@

The methodology proposed in this work relies heavily on the use of RKHS's, so before diving into it we will briefly describe the main characteristics of these spaces. In what follows, suppose \(X=X(t)\) is an \(L^2\)-process with trajectories in \(L^2[0,1]\), and further assume that its mean function \(m(t)=\mathbb E[X(t)]\) and its covariance function \(K(t, s)= \mathbb E[(X(t) - m(t))(X(s) - m(s))]\) are both continuous. To construct the RKHS associated with the covariance function, we start by defining the functional space \(\mathcal H_0(K)\) of all finite linear combinations of evaluations of \(K\), that is,
\[
\mathcal H_0(K) = \left\{ f \in L^2[0,1]: f(\cdot) = \sum_{i=1}^p a_i K(t_i, \cdot), \ p \in \N, \ a_i \in \R, \ t_i \in [0, 1] \right\}.
\]
This space is endowed with the inner product
\[
\langle f, g\rangle_K = \sum_{i, j} a_i b_j K(t_i, s_j),
\]
given that \(f(\cdot)=\sum_i a_i K(t_i, \cdot) \) and \(g(\cdot)=\sum_j b_j K(s_j, \cdot)\). Lastly, we define \(\mathcal H(K)\) to be the completion of \(\mathcal H_0(K)\) under the norm induced by the scalar product defined above.

\begin{comment}
  \(K\) is positive and symmetric. We will refer to it as a \textit{kernel}.
\end{comment}

\begin{comment}
  Reproducing property? Convergence entails uniform convergence?
\end{comment}

As we can see, the covariance function plays a crucial role in characterizing the RKHS. A compact operator closely related to this covariance function is the so-called covariance operator, i.e.:
\[
\mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds, \quad f \in L^2[0, 1].
\]
This operator is especially relevant due to the \textit{Karhunen-Loève} theorem \incomment{cita}, which asserts that \(X\) can be decomposed as
\begin{equation}\label{eq:karhunen-loeve}
X(t) = \sum_{j=1}^\infty \xi_j \phi_j(t),
\end{equation}
where \(\phi_j\) are the (orthonormal) eigenfunctions of the covariance operator \(\mathcal K\) and \(\xi_j = \int X\phi_j\) are uncorrelated zero-mean random variables with variance equal to \(\lambda_j\), the eigenvalue associated with \(\phi_j\). This expansion is the main justification of many FDA techniques, such as functional principal component analysis \incomment{cita}.

@@

\begin{enumerate}
  \item Via \textit{Loeve's isometry}, one can establish a congruence between \(\mathcal H(K)\) and the linear span of the centered process \(X - m\) in the space of all random variables with finite second moment, \(L^2(\Omega)\) \incomment{cita}. This isometry is essentially the completion of the correspondence
  \begin{equation}\label{eq:loeve}
  \sum_{i=1}^p a_i (X(t_i) - m(t_i)) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot).
\end{equation}
  \item The space \(\mathcal H(K)\) can also be defined as the image of the square root of the covariance operator, i.e.:
  \begin{equation}\label{eq:rkhs-square-root}
  \mathcal H(K) = \mathcal K^{1/2}(L^2[0, 1]).
\end{equation}

  \item Lastly, starting from the above definition and combining expression~\eqref{eq:karhunen-loeve} and the representation of \(K\) given by \textit{Mercer's theorem} \incomment{cita}, we can also express
  \begin{equation}\label{rkhs-sum-lambda}
    \mathcal H(K) = \left\{f \in L^2[0, 1]: \sum_{j=1}^\infty \frac{\langle f, \phi_j\rangle}{\lambda_j} < \infty \right\},
  \end{equation}
  with the corresponding inner product
  \[
  \langle f, g\rangle_K = \sum_{j=1}^\infty \frac{\langle f, \phi_j\rangle\langle g, \phi_j\rangle}{\lambda_j}.
  \]
Note that, since the spectral theorem for compact operators tells us that \(\lambda_j \to 0\) \incomment{cita}, this definition highlights the fact that functions in \(\mathcal H(K)\) are smooth, in the sense that their components in an orthonormal basis need to vanish quickly \incomment{and?}.
\end{enumerate}

\begin{comment}
  Trajectories not in RKHS
\end{comment}

\begin{comment}
  Small RKHS norm implies small variations and hence a higher degree of smoothness. Specifically, \(\|f\|_K\) controls how fast the function varies over the input space with respect to the geometry defined by the kernel:
  \[
  |f(t) - f(s)| = |\langle f, K_t - K_s\rangle_K| \leq \|f\|_K \|K_t - K_s\|_K.
  \]
\end{comment}

@@

@@

Note that starting from a probability distribution \(\mathbb{P}_0\) on \(\mathcal H_0(K)\) we can obtain a probability distribution \(\mathbb{P}\) on \(\mathcal H(K)\) simply by defining \(\mathbb{P}(B) = \mathbb{P}_0(B\cap \mathcal H_0(K))\) for all Borel sets \(B\). The following result proves that any such extension is indeed unique.

\begin{prop} Let \(H\) be a separable Hilbert space and consider a dense subset \(H_0\subseteq H\). Then, any probability distribution on \((H, \mathcal{B}(H))\) is uniquely determined by its restriction to \((H_0, \mathcal B(H_0))\).
\end{prop}

\begin{comment}
  ¿Es necesario especificar que \(\mathcal B\) es la sigma álgebra de Borel?
\end{comment}

\begin{comment}
    ¿La prueba va en el cuerpo del artículo, o en un anexo/material suplementario?
\end{comment}

@@

First, note that Theorem 7.1.2 in \citet[p.~177]{hsing2015theoretical} establishes that the distribution of any random variable \(X: (\Omega, \mathcal A, \mathbb{P})\to (H, \mathcal B(H))\) is uniquely determined by the distribution of its random projections \(\langle X, h\rangle_H\) for all \(h \in H\).
Consider two random variables \(\alpha\) and \(\beta\), taking values in \(H\) and equally distributed on \(H_0\), and let \(h \in H\). Since \(H_0\) is dense in \(H\), there is a sequence \(\{h_n\}\subseteq H_0\) such that \(\|h_n - h\|_H \to 0\), and necessarily \(\langle \alpha, h_n\rangle_H \equiv \langle \beta, h_n\rangle_H\) for all \(n\), where the equality is in distribution.

Observe now that \(\langle \alpha, h_n\rangle_H \overset{P}{\to} \langle \alpha, h\rangle_H\), since
\[
\forall \epsilon > 0 \quad \mathbb{P}\{|\langle \alpha, h_n - h\rangle_H| > \epsilon\} \leq \mathbb{P}\{\|\alpha\|_H \|h_n - h\|_H > \epsilon\} \to 0,
\]
and equivalently \(\langle \beta, h_n \rangle_H \overset{P}{\to} \langle \beta, h\rangle_H\), so that \(\langle \alpha, h \rangle_H \equiv \langle \beta,h\rangle_H\). As this is valid for every \(h \in H\), we can conclude that \(\alpha \equiv \beta\) on \(H\).
\end{proof}

\begin{comment}
  Note that in an RKHS the evaluation functional is continuous, so a small perturbation \(f\mapsto f + \delta f\) results in a small change in \(f(x)\) for all \(x\). It makes sense to choose the parameter in \(\mathcal H(K)\). Also smooth functions and linear: the model is \textit{linear} because of the correspondence between the RKHS and the linear span of the process (Loeve's isometry).
\end{comment}

\begin{comment}
  Overview of MCMC methods.
\end{comment}

@@

Berrendero et al. (2020).

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supplementary Material, if any, should   %%
%% be provided in {supplement} environment  %%
%% with title and short description.        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supplement}
\stitle{Title of Supplement A}
\sdescription{Short description of Supplement A.}
\end{supplement}

>>>>

\newcommand{\N}{\mathbb{N}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}

%% Scalar product
\newcommand\dotprod[2]{\left\langle #1,\, #2 \right\rangle}

@@

%% Inline comments
\newcommand\incomment[1]{\color{red}[\textit{#1}]\color{black}}
\endlocaldefs

@@

\begin{comment}
  - Grollemund et al.

  - FDA (Cuevas et al. 2013)

  - L2 model and finite model as a particular case of RKHS model. Define dot-product in L2.
\end{comment}

@@

\begin{comment}
  Es posible que esta sección sea demasiado larga...?
\end{comment}

The methodology proposed in this work relies heavily on the use of RKHS's, so before diving into it we will briefly describe the main characteristics of these spaces. In what follows, suppose \(X=X(t)\) is an \(L^2\)-process with trajectories in \(L^2[0,1]\), and further assume that its mean function \(m(t)=\mathbb E[X(t)]\) and its covariance function \(K(t, s)= \mathbb E[(X(t) - m(t))(X(s) - m(s))]\) are both continuous. To construct the RKHS associated with the covariance function, we start by defining the functional space \(\mathcal H_0(K)\) of all finite linear combinations of evaluations of \(K\), that is,
\[
\mathcal H_0(K) = \left\{ f \in L^2[0,1]: f(\cdot) = \sum_{i=1}^p a_i K(t_i, \cdot), \ p \in \N, \ a_i \in \R, \ t_i \in [0, 1] \right\}.
\]
This space is endowed with the inner product
\[
\dotprod{f}{g}_K = \sum_{i, j} a_i b_j K(t_i, s_j),
\]
given that \(f(\cdot)=\sum_i a_i K(t_i, \cdot) \) and \(g(\cdot)=\sum_j b_j K(s_j, \cdot)\). Lastly, we define \(\mathcal H(K)\) to be the completion of \(\mathcal H_0(K)\) under the norm induced by the scalar product defined above. As it turns out, functions in this space verify the \textit{reproducing property}, thus giving it its name:
\begin{equation}\label{eq:reproducing-property}
  \dotprod{K(t, \cdot)}{f}_K = f(t), \quad \text{for all } f \in \mathcal H(K), \ t \in [0, 1].
\end{equation}

As we can see, the covariance function (sometimes referred to as the \textit{kernel}) plays a crucial role in characterizing the RKHS. A linear compact operator closely related to this covariance function is the so-called covariance operator, i.e.:
\[
\mathcal Kf(\cdot) = \int_0^1 K(s, \cdot)f(s)\, ds, \quad f \in L^2[0, 1].
\]
This operator is especially relevant due to the \textit{Karhunen-Loève} theorem \incomment{cita}, which asserts that a centered process \(X\) can be decomposed as
\begin{equation}\label{eq:karhunen-loeve}
X(t) = \sum_{j=1}^\infty \xi_j \phi_j(t),
\end{equation}
where \(\phi_j\) are the (orthonormal) eigenfunctions of its covariance operator \(\mathcal K\) and \(\xi_j = \int X\phi_j\) are uncorrelated zero-mean random variables with variance equal to \(\lambda_j\), the eigenvalue associated with \(\phi_j\). This expansion is the main justification of many FDA techniques, such as functional principal component analysis (FPCA) (see for example \incomment{cita}).

@@

\begin{enumerate}
  \item Via \textit{Loeve's isometry}, one can establish a congruence between \(\mathcal H(K)\) and the linear span of the centered process, \(\mathcal L(X)\), in the space of all random variables with finite second moment, \(L^2(\Omega)\) \incomment{cita}. This isometry is essentially the completion of the correspondence
  \[
  \sum_{i=1}^p a_i (X(t_i) - m(t_i)) \longleftrightarrow \sum_{i=1}^p a_i K(t_i, \cdot),
\]
and can be formally defined as (see \incomment{cita})
\begin{equation}\label{eq:loeves-isometry}
  \Psi_X(U)(t) = \E[U(X(t) - m(t))], \quad U \in \mathcal L(X).
\end{equation}
  \item The space \(\mathcal H(K)\) can also be defined as the image of the square root of the covariance operator, i.e.:
  \begin{equation}\label{eq:rkhs-square-root}
  \mathcal H(K) = \mathcal K^{1/2}(L^2[0, 1]).
\end{equation}
In this case, the inner product can be expressed as
\[
\dotprod{f}{g}_K = \dotprod{\mathcal K^{1/2}(f)}{\mathcal K^{1/2}(g)}.
\]

  \item Lastly, starting from the above definition and combining expression~\eqref{eq:karhunen-loeve} and the representation of \(K\) given by \textit{Mercer's theorem} \incomment{cita}, we can also express
  \begin{equation}\label{rkhs-sum-lambda}
    \mathcal H(K) = \left\{f \in L^2[0, 1]: \sum_{j=1}^\infty \frac{\dotprod{f}{\phi_j}^2}{\lambda_j} < \infty \right\},
  \end{equation}
  with the corresponding inner product
  \[
  \dotprod{f}{g}_K = \sum_{j=1}^\infty \frac{\dotprod{f}
  {\phi_j}\dotprod{g}{\phi_j}}{\lambda_j}.
  \]
Note that, since the spectral theorem for compact operators tells us that \(\lambda_j \to 0\) \incomment{cita}, this definition highlights the fact that functions in \(\mathcal H(K)\) are smooth, in the sense that their components in an orthonormal basis need to vanish quickly.
\end{enumerate}

Despite the close connection between the process \(X\) and the space \(\mathcal H(K)\), special care must be taken when dealing with concrete realizations of the process, since in general the trajectories do not belong to the RKHS with probability one (see \incomment{cita}). As a consequence, the expression \(\dotprod{x}{f}_K\) is ill-defined for \(x\) a realization of \(X\). However, following Parzen's approach \incomment{cita}\ we can leverage Loève's isometry and identify \(\dotprod{x}{f}_K \) with \( \Psi_X^{-1}(f)(\omega)\), for \(x=X(\omega)\) and \(f\in \mathcal H(K)\). This notation often proves to be useful and convenient.

@@

The classical functional linear regression model, firstly introduced in \incomment{cita Ramsay \& Silverman}, can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^n\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-model}
Y = \dotprod{X}{\beta} + \epsilon = \int_0^1 X(t)\beta(t)\, dt + \epsilon.
\end{equation}
In this case, the main parameter \(\beta=\beta(t)\) is a member of the infinte-dimensional space \(L^2[0, 1]\). Aside from the fact that functional (i.e. infinite-dimensional) optimization is a hard problem in and of itself \incomment{cita?}, there are a number of more subtle inconveniences with this model. For instance, \(L^2[0, 1]\) is a big space that contains many non-smooth or ill-behaved functions, but incidentally the model given by~\eqref{eq:l2-model} does not include a finite-dimensional model based on a linear combination of projections of \(X\) as a particular case (see \incomment{Berrendero et al 2020, on a general definition of the flm.}).

It turns out that a natural alternative to the \(L^2\) mode is the RKHS model..... Not only simpler, but also smoother and more manageable functions. The model remains linear because of the correspondence between the RKHS and the linear span of the process (Loève’s isometry).

\begin{comment}
  - The RKHS model includes both L2 model and finite model. Cite Berrendero et al 2020.
  - Parametric approach with \(\beta \in H_0(K)\) (further constrain the parameter to this subset of the RKHS, which is dense by construction). Then we can impose a prior on the finite-dimensional parameter space.
\end{comment}

@@

Note that starting from a probability distribution \(\mathbb{P}_0\) on \(\mathcal H_0(K)\) we can obtain a probability distribution \(\mathbb{P}\) on \(\mathcal H(K)\) simply by defining \(\mathbb{P}(B) = \mathbb{P}_0(B\cap \mathcal H_0(K))\) for all Borel sets \(B \in \mathcal B(\mathcal H(K))\). The following result proves that any such extension is indeed unique.

\begin{prop} Let \(H\) be a separable Hilbert space and consider a dense subset \(H_0\subseteq H\). Then, any probability distribution on \((H, \mathcal{B}(H))\) is uniquely determined by its restriction to \((H_0, \mathcal B(H_0))\).
\end{prop}
\begin{comment}
    ¿La prueba va en el cuerpo del artículo, o en un anexo/material suplementario?
\end{comment}

@@

First, note that Theorem 7.1.2 in \citet[p.~177]{hsing2015theoretical} establishes that the distribution of any random variable \(X: (\Omega, \mathcal A, \mathbb{P})\to (H, \mathcal B(H))\) is uniquely determined by the distribution of its random projections \(\dotprod{X}{h}_H\) for all \(h \in H\).
Consider two random variables \(\alpha\) and \(\beta\), taking values in \(H\) and equally distributed on \(H_0\), and let \(h \in H\). Since \(H_0\) is dense in \(H\), there is a sequence \(\{h_n\}\subseteq H_0\) such that \(\|h_n - h\|_H \to 0\), and necessarily \(\dotprod{\alpha}{h_n}_H \equiv \dotprod{\beta}{h_n}_H\) for all \(n\), where the equality is in distribution.

Observe now that \(\dotprod{\alpha}{h_n}_H \overset{P}{\to} \dotprod{\alpha}{h}_H\), since
\[
\forall \epsilon > 0 \quad \mathbb{P}\{|\dotprod{\alpha}{h_n - h}_H| > \epsilon\} \leq \mathbb{P}\{\|\alpha\|_H \|h_n - h\|_H > \epsilon\} \to 0,
\]
and equivalently \(\dotprod{\beta}{h_n}_H \overset{P}{\to} \dotprod{\beta}{h}_H\), so that \(\dotprod{\alpha}{h}_H \equiv \dotprod{\beta}{h}_H\). As this is valid for every \(h \in H\), we can conclude that \(\alpha \equiv \beta\) on \(H\).
\end{proof}

\subsubsection{Overview of MCMC methods}

@@

\begin{comment}
  - Estimators derived.
\end{comment}

@@

In this case the model/functional equation in $L^2$ is....... and the same problems apply, plus the MLE existence problem. An RKHS aproach is given by:.......

 and this solves the problems (see Bueno-Larraz 2021)

\begin{comment}
  - Estimators derived.
\end{comment}

\subsection{Model choice and validation}

\begin{comment}
  - Choice of p, MLE, BIC, ...

  - Posterior predictive checks, ¿Bayesian p-values?, ...
\end{comment}

@@

\begin{comment}
  Not only MSE/Acc but also credible intervals, points-of-impact model --> variable selection included.
\end{comment}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supplementary Material, if any, should   %%
%% be provided in {supplement} environment  %%
%% with title and short description.        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supplement}
\stitle{Title of Supplement A}
\sdescription{Short description of Supplement A, with DOI and/or links to additional material (code, ...)}
\end{supplement}
